DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-06-23 manuscript/manuscript.Rmd
supplementary files for biorxiv submission

At the time of writing of this manuscript 
(`r format(Sys.time(), "%b %d, %Y")`),
DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.

@@

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species. 
The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).

@@

Age data from source chronograms was congruified to OpenTree's topology (Fig. \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table 2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S4).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

@@

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9). The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1 and S4).

>>>>

At the time of writing of this manuscript (Jun 22, 2022),
<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.

@@

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).

@@

Age data from source chronograms was congruified to OpenTree's topology (Fig. \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

@@

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20). The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1, S5).

### 2022-06-23 manuscript/manuscript.tex
supplementary files for biorxiv submission

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4924}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
For example, the chronograms from Burns et al. (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S4).

@@

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9). The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al. (2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1 and S4).

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4926}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
For example, the chronograms from Burns et al. (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).

@@

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20). The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al. (2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs. S1, S5).
